---
title: "Cyclistic Bike-Share"
subtitle: "How annual members and casual riders use bikes differently"
author: "Eduardo Escudero"
format:
  revealjs:
    theme: [simple, reveal_refined.scss]
    slide-level: 0
    slide-number: c/t
    controls: true
    progress: true
    history: false
    hash: false
    navigation-mode: linear
    transition: fade
    center: false
    embed-resources: true
    footer: "Cyclistic Case Study"
    width: 1600
    height: 900
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(lubridate)
library(scales)

options(scipen = 999)
knitr::opts_chunk$set(fig.align = "center", out.width = "90%")

path <- "C:/Users/G18333/Documents/Eduardo Escudero/Cursos/Coursera/Google data analytics/Capstone/Data_capstone"

user_palette <- c(casual = "#E67E22", member = "#1F78B4")
user_labels <- c(casual = "Casual", member = "Member")
weekday_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
weekday_short <- c("Monday" = "Mon", "Tuesday" = "Tue", "Wednesday" = "Wed", "Thursday" = "Thu", "Friday" = "Fri", "Saturday" = "Sat", "Sunday" = "Sun")
month_lookup <- c(
  "01" = "January", "02" = "February", "03" = "March", "04" = "April",
  "05" = "May", "06" = "June", "07" = "July", "08" = "August",
  "09" = "September", "10" = "October", "11" = "November", "12" = "December"
)
month_short <- c(
  "January" = "Jan", "February" = "Feb", "March" = "Mar", "April" = "Apr",
  "May" = "May", "June" = "Jun", "July" = "Jul", "August" = "Aug",
  "September" = "Sep", "October" = "Oct", "November" = "Nov", "December" = "Dec"
)

plot_theme <- theme_minimal(base_size = 18) +
  theme(
    plot.title.position = "plot",
    plot.caption.position = "plot",
    plot.title = element_text(face = "bold", size = 19, colour = "#183247", margin = margin(b = 6)),
    plot.subtitle = element_text(size = 11.5, colour = "#5B616B", margin = margin(b = 10)),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title = element_text(face = "bold", size = 12.5),
    axis.text = element_text(size = 10.5, colour = "#304454"),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(size = 10.5),
    plot.margin = margin(10, 26, 10, 10)
  )

ggplot2::theme_set(plot_theme)

files <- list.files(
  path = path,
  pattern = "^\\d{6}-divvy-tripdata\\.csv$",
  full.names = TRUE
) |> sort()

if (length(files) == 0) {
  stop("No files matching YYYYMM-divvy-tripdata.csv were found in: ", path)
}

divvy_all <- purrr::map_dfr(files, function(f) {
  m <- stringr::str_match(basename(f), "^(\\d{4})(\\d{2})-divvy-tripdata\\.csv$")
  readr::read_csv(f, show_col_types = FALSE) %>%
    mutate(
      year = as.integer(m[2]),
      month = as.integer(m[3]),
      month_name = factor(month_lookup[m[3]], levels = unname(month_lookup), ordered = TRUE)
    )
})

divvy_all <- divvy_all %>%
  mutate(
    started_at = ymd_hms(started_at),
    ended_at = ymd_hms(ended_at),
    start_date = as.Date(started_at),
    weekday_num = wday(start_date, week_start = 1),
    weekday = factor(weekday_levels[weekday_num], levels = weekday_levels, ordered = TRUE),
    duration_min = as.numeric(difftime(ended_at, started_at, units = "mins")),
    duration_sec = as.numeric(difftime(ended_at, started_at, units = "secs"))
  )

divvy_filter_base <- divvy_all %>%
  filter(duration_sec > 0, !is.na(end_lat), !is.na(end_lng))

cutoff_995 <- quantile(divvy_filter_base$duration_sec, 0.995, na.rm = TRUE)

divvy_filter <- divvy_filter_base %>%
  filter(duration_sec <= cutoff_995) %>%
  mutate(start_hour = hour(started_at))

removed_pct <- (nrow(divvy_all) - nrow(divvy_filter)) / nrow(divvy_all)

avg_ride_by_user <- divvy_filter %>%
  group_by(member_casual) %>%
  summarise(avg_ride_length_min = round(mean(duration_min, na.rm = TRUE), 2), .groups = "drop")

rides_by_user_weekday <- divvy_filter %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), .groups = "drop") %>%
  group_by(member_casual) %>%
  mutate(pct_within_user_type = number_of_rides / sum(number_of_rides)) %>%
  ungroup()

avg_ride_by_user_weekday <- divvy_filter %>%
  group_by(member_casual, weekday) %>%
  summarise(avg_ride_length_min = mean(duration_min, na.rm = TRUE), .groups = "drop")

rides_by_hour <- divvy_filter %>%
  group_by(member_casual, start_hour) %>%
  summarise(number_of_rides = n(), .groups = "drop") %>%
  group_by(member_casual) %>%
  mutate(pct_within_user_type = number_of_rides / sum(number_of_rides)) %>%
  ungroup()

rides_by_month <- divvy_filter %>%
  group_by(member_casual, month_name) %>%
  summarise(number_of_rides = n(), .groups = "drop")

seasonality_summary <- rides_by_month %>%
  group_by(member_casual) %>%
  summarise(
    low_month = as.character(month_name[which.min(number_of_rides)]),
    high_month = as.character(month_name[which.max(number_of_rides)]),
    low_rides = min(number_of_rides),
    high_rides = max(number_of_rides),
    peak_to_trough = round(high_rides / low_rides, 1),
    .groups = "drop"
  )

summer_gap <- rides_by_month %>%
  filter(month_name == "August") %>%
  summarise(ratio = round(number_of_rides[member_casual == "casual"] / number_of_rides[member_casual == "member"], 2)) %>%
  pull(ratio)

winter_gap <- rides_by_month %>%
  filter(month_name == "January") %>%
  summarise(ratio = round(number_of_rides[member_casual == "casual"] / number_of_rides[member_casual == "member"], 2)) %>%
  pull(ratio)

weekend_share <- rides_by_user_weekday %>%
  mutate(day_type = if_else(weekday %in% c("Saturday", "Sunday"), "Weekend", "Weekday")) %>%
  group_by(member_casual, day_type) %>%
  summarise(pct = sum(pct_within_user_type), .groups = "drop")

peak_hour <- rides_by_hour %>%
  group_by(member_casual) %>%
  slice_max(order_by = pct_within_user_type, n = 1, with_ties = FALSE) %>%
  ungroup()

casual_avg <- avg_ride_by_user$avg_ride_length_min[avg_ride_by_user$member_casual == "casual"]
member_avg <- avg_ride_by_user$avg_ride_length_min[avg_ride_by_user$member_casual == "member"]
casual_weekend <- weekend_share$pct[weekend_share$member_casual == "casual" & weekend_share$day_type == "Weekend"]
member_weekend <- weekend_share$pct[weekend_share$member_casual == "member" & weekend_share$day_type == "Weekend"]
casual_peak_hour <- peak_hour$start_hour[peak_hour$member_casual == "casual"]
member_peak_hour <- peak_hour$start_hour[peak_hour$member_casual == "member"]
```

## Cyclistic Bike-Share

::: hero-box
**Casual riders behave more like leisure users. Members behave more like routine users.**
:::

::: eyebrow
Executive snapshot
:::

:::::::::::: summary-grid
::::: summary-card
::: card-label
Ride length
:::

::: big-number
`r casual_avg` min
:::

Casual average trip length versus **`r member_avg` min** for members.
:::::

::::: summary-card
::: card-label
Weekend weight
:::

::: big-number
`r percent(casual_weekend, accuracy = 0.1)`
:::

Weekend share for casual riders versus **`r percent(member_weekend, accuracy = 0.1)`** for members.
:::::

::::: summary-card
::: card-label
Seasonality
:::

::: big-number
`r seasonality_summary$peak_to_trough[seasonality_summary$member_casual == "casual"]`x
:::

Casual peak-to-trough monthly swing versus **`r seasonality_summary$peak_to_trough[seasonality_summary$member_casual == "member"]`x** for members.
:::::
::::::::::::

::: mini-note
This deck keeps the structure flat and linear to avoid navigation issues while still adding more visual polish.
:::

------------------------------------------------------------------------

## Agenda

::: chip-row
[Business question]{.chip} [Data scope]{.chip} [Ride length]{.chip} [Weekday mix]{.chip} [Time of day]{.chip} [Monthly pattern]{.chip} [Recommendations]{.chip}
:::

::: takeaway-box
**Presentation logic:** start with the cleaned data, then compare trip duration, weekly timing, hourly timing, and monthly demand.
:::

------------------------------------------------------------------------

## Business question

::: hero-box
**How do annual members and casual riders use Cyclistic bikes differently?**
:::

-   Members are more profitable than casual riders.
-   Marketing needs evidence on when, how, and how long each group rides.
-   The business goal is to identify where casual behavior looks most convertible into membership.

::: takeaway-box
**Takeaway:** the analysis focuses on behavioral differences that can support membership conversion strategy.
:::

------------------------------------------------------------------------

## Data scope and filtering

::::::::::::::: stat-grid
::::: stat-card
::: card-label
Monthly files
:::

::: big-number
12
:::

Trip-level historical data for 2025.
:::::

::::: stat-card
::: card-label
Trips loaded
:::

::: big-number
`r format(nrow(divvy_all), big.mark = ",")`
:::

Raw observations before filtering.
:::::

::::: stat-card
::: card-label
Trips kept
:::

::: big-number
`r format(nrow(divvy_filter), big.mark = ",")`
:::

Cleaned sample used in the analysis.
:::::

::::: stat-card
::: card-label
Trips removed
:::

::: big-number
`r percent(removed_pct, accuracy = 0.1)`
:::

Invalid or extreme observations excluded.
:::::
:::::::::::::::

Filtering rules: - Remove non-positive duration. - Remove trips with missing end coordinates. - Remove rides above the 99.5th percentile of duration.

::: takeaway-box
**Takeaway:** findings are based on a cleaned sample of plausible rides, not on raw operational logs.
:::

------------------------------------------------------------------------

## Why this comparison matters

::::::::: summary-grid
:::: summary-card
::: card-label
Casual riders
:::

Longer trips, heavier weekend usage, stronger seasonal spikes.
::::

:::: summary-card
::: card-label
Members
:::

Shorter trips, more weekday concentration, steadier demand over time.
::::

:::: summary-card
::: card-label
Implication
:::

Conversion messaging should emphasize convenience, routine value, and repeated use.
::::
:::::::::

::: takeaway-box
**Takeaway:** the strongest evidence points to different trip purposes, not just different ride volumes.
:::

------------------------------------------------------------------------

## Key finding 1: casual riders take longer trips

```{r}
#| fig-height: 5.7
#| fig-width: 11.2
#| fig-align: center

avg_ride_by_user %>%
  ggplot(aes(x = member_casual, y = avg_ride_length_min, fill = member_casual)) +
  geom_col(width = 0.62) +
  geom_text(aes(label = round(avg_ride_length_min, 2)), vjust = -0.35, size = 5.7, fontface = "bold") +
  scale_fill_manual(values = user_palette, labels = user_labels) +
  scale_x_discrete(labels = user_labels) +
  scale_y_continuous(
    limits = c(0, max(avg_ride_by_user$avg_ride_length_min) * 1.22),
    expand = expansion(mult = c(0, 0.02))
  ) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Average ride length by rider type",
    subtitle = "Casual riders spend more time per trip even after filtering invalid and extreme rides",
    x = NULL,
    y = "Average ride length (minutes)"
  ) +
  theme(legend.position = "none")
```

::: takeaway-box
**Takeaway:** casual riders spend more time per trip, consistent with more leisure-oriented usage.
:::

------------------------------------------------------------------------

## Key finding 2: the weekday mix is different

```{r}
#| fig-height: 5.8
#| fig-width: 11.4
#| fig-align: center

rides_by_user_weekday %>%
  ggplot(aes(x = weekday, y = pct_within_user_type, fill = member_casual)) +
  geom_col(position = position_dodge(width = 0.76), width = 0.68) +
  scale_fill_manual(values = user_palette, labels = user_labels) +
  scale_x_discrete(labels = weekday_short) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    title = "Share of rides by weekday",
    subtitle = "Percent of rides within each rider type",
    x = NULL,
    y = "Share of rides",
    fill = NULL
  )
```

::: takeaway-box
**Takeaway:** casual riders are more concentrated on weekends, while members show a stronger weekday profile.
:::

------------------------------------------------------------------------

## Key finding 2b: casual riders also ride longer on weekends

```{r}
#| fig-height: 5.8
#| fig-width: 11.4
#| fig-align: center

avg_ride_by_user_weekday %>%
  ggplot(aes(x = weekday, y = avg_ride_length_min, color = member_casual, group = member_casual)) +
  geom_line(linewidth = 1.25) +
  geom_point(size = 3.2) +
  scale_color_manual(values = user_palette, labels = user_labels) +
  scale_x_discrete(labels = weekday_short) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.09))) +
  labs(
    title = "Average ride length across the week",
    subtitle = "Casual riders stay above members every day and peak on weekends",
    x = NULL,
    y = "Average ride length (minutes)",
    color = NULL
  )
```

::: takeaway-box
**Takeaway:** the difference is not only when each group rides, but also how they ride across the week.
:::

------------------------------------------------------------------------

## Key finding 3: time of day separates the groups

```{r}
#| fig-height: 5.8
#| fig-width: 11.4
#| fig-align: center

rides_by_hour %>%
  ggplot(aes(x = start_hour, y = pct_within_user_type, color = member_casual)) +
  geom_line(linewidth = 1.25) +
  scale_color_manual(values = user_palette, labels = user_labels) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0.02, 0.08))
  ) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 16, 20, 23)) +
  labs(
    title = "Hourly distribution of rides",
    subtitle = glue::glue("Peak member hour: {member_peak_hour}:00 | Peak casual hour: {casual_peak_hour}:00"),
    x = "Hour of day",
    y = "Share of rides",
    color = NULL
  )
```

::: takeaway-box
**Takeaway:** members look more routine-oriented, while casual riders appear more tied to free time than fixed commuting windows.
:::

------------------------------------------------------------------------

## Key finding 4: both groups are seasonal, but not equally

```{r}
#| fig-height: 5.9
#| fig-width: 11.4
#| fig-align: center

rides_by_month %>%
  ggplot(aes(x = month_name, y = number_of_rides, color = member_casual, group = member_casual)) +
  geom_line(linewidth = 1.25) +
  geom_point(size = 3) +
  scale_color_manual(values = user_palette, labels = user_labels) +
  scale_x_discrete(labels = month_short) +
  scale_y_continuous(
    labels = label_number(scale_cut = cut_short_scale()),
    expand = expansion(mult = c(0.02, 0.08))
  ) +
  labs(
    title = "Monthly ride volume",
    subtitle = "Members lead every month, but casual demand swings much more strongly",
    x = NULL,
    y = "Number of rides",
    color = NULL
  )
```

-   Both groups rise from spring into summer and decline in winter.
-   Casual riders swing about **`r seasonality_summary$peak_to_trough[seasonality_summary$member_casual == "casual"]`x** from `r seasonality_summary$low_month[seasonality_summary$member_casual == "casual"]` to `r seasonality_summary$high_month[seasonality_summary$member_casual == "casual"]`.
-   Members also vary seasonally, but less sharply: about **`r seasonality_summary$peak_to_trough[seasonality_summary$member_casual == "member"]`x**.
-   The casual/member ratio rises from **`r winter_gap`** in January to **`r summer_gap`** in August.

::: takeaway-box
**Takeaway:** seasonality affects both groups, but it amplifies casual demand much more strongly.
:::

------------------------------------------------------------------------

## Recommendations

:::::::::::: summary-grid
::::::::::: summary-card
:::::::::: card-label
1.  Convert weekend casuals

    </div>

    Focus membership offers on high-leisure periods, especially Saturday and Sunday trips.

:::

::::::::: summary-card
:::::::: card-label
2.  Sell convenience and routine

    </div>

    Message membership as a better fit for repeated, shorter, more frequent usage.

:::

::::::: summary-card
:::::: card-label
3.  Time campaigns with demand peaks

    </div>

    Prioritize months and hours when casual demand accelerates and repeat behavior is more likely.

:::

:::

::: takeaway-box
**Takeaway:** the best conversion opportunities appear where casual demand is high, recurring, and seasonally amplified.
:::

------------------------------------------------------------------------

## Bottom line

::: hero-box
**Casual riders look more leisure-oriented. Members look more routine-oriented.**
:::

The strongest differences are visible in: - trip duration - weekday distribution - time-of-day behavior - the strength of seasonal demand

::: mini-note
This version keeps the slide structure flat on purpose. If it renders cleanly, the next risky feature to test would be a multi-column content slide.
:::
::::::
:::::::
::::::::
:::::::::
::::::::::
:::::::::::
::::::::::::
